The unequal distribution of access to education and transport: insights from

### 2016-06-02 paper.Rmd
Include formatted table
Latest paper update (from Ana)

In many large Latin American cities such as São Paulo, growing social
and economic inequalities are embedded through unfair education and
transport systems. Inequalities in educational and transport
infrastructure are mutually reinforcing: the right to mobility is
intrinsically linked to the right to education. This is manifested by
the overlap between recent protests against unwanted changes to public
education and the social movements contesting increases in public
transport fares. Another manifestation is to be found in the concept of
school accessibility. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied to São Paulo. By capturing both the unequal distribution of schools
and transport services across space, the index allows embedded
inequalities to be better understood and, with political will,
contested. Our index combines information on the spatial distribution of
adolescents, the location of existing schools, the travel patterns of students, and the travel infrastructure serving the school catchment area into a single measure. The index is used to measure school accessibility locally across São Paulo, using
data sources from Population and School Censuses, commonly available in
Latin American cities. The results illustrate how existing inequalities
are amplified by variable accessibility to schools across income groups
and geographical space. We conclude that extending the concept of local
accessibility indicators to education can help to both contest and
constructively tackle embedded social inequalities.

@@

Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the São Paulo Metropolitan Region gained access to a public transport subsidy in 2015. However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016. The change in policy was estimated to affect over 300,000 students, many whom would be placed in schools far from their homes. As a response, in the following months students occupied over 200 schools. Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (The New York Times, 2015).

@@

<!-- (INCLUDE SOMETHING ABOUT CONTESTATION) -->

The new index contributes to the measurement of school accessibility. It is inspired by the index developed by Shen (1998) in the context of job accessibility. The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area. Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available. As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas. The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas. The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space.

HERE REVIEW PREVIOUS SCHOOL ACCESSIBILITY MEASURES AND EXPLAIN THEIR LIMITATIONS VIS-A-VIS OUR INDEX
--Lit review

The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay. This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].

References transport and social exclusion
[@jones2012] provides a general framework to understand transport-related social exclusion in emerging economies

Evidence on access inequalities in the Brazilian context [@gomide2006]

Active travel children and adolescents in the SPMR [@de2015]

Previous literature for consolidated has found that school accessibility has a significant and positive impact on graduation from upper secondary schools [@falch2013], and on post-compulsory education enrolment [@dickerson2012]. For emerging economies, [@asahi2014] finds that improved metro accessibility had a negative impact on student's performance as measured by test scores.  
...

@@

```{r, echo=FALSE, fig.cap="The home location of students attending public secondary schools students in São Paulo.", out.width="70%"}
knitr::include_graphics(c("med_pu.png"))
```

```{r, echo=FALSE, fig.cap="The home location of students attending private secondary schools students in São Paulo.", out.width="70%"}
knitr::include_graphics(c("med_pr.png"))
```

<!-- ![](med_pr.png) -->
<!-- ![Public secondary schools students home location](med_pu.png) -->

@@

![Enrollment in public secondary schools](enrol.png)

@@

```{r table-placeholder, echo=FALSE}

```

@@

# Discussion

@@

>>>>

In many large Latin American urban areas such as the São Paulo Metropolitan Region (SPMR), growing social
and economic inequalities are embedded through unfair education and
transport systems. Inequalities in educational and transport
infrastructure are mutually reinforcing: the right to mobility is
intrinsically linked to the right to education. This is manifested by
the overlap between recent protests against unwanted changes to public
education and the social movements contesting increases in public
transport fares. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied to the SPMR. Our index combines information on the spatial distribution of
adolescents, the location of existing schools, and the travel infrastructure serving the school catchment area into a single measure. The index is used to measure school accessibility locally across 633 areas within the SPMR. The results illustrate how existing inequalities
are amplified by variable accessibility to schools across income groups
and geographical space.

@@

The São Paulo Metropolitan Region (SPMR) presents an excellent opportunity to measure the extent of educational and transport infrastructure, and to study the impact of public policies on different socio-economic groups. Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015. However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016. The change in policy was estimated to affect over 300,000 students, many whom would be placed in schools far from their homes. As a response, in the following months students occupied over 200 schools. Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (The New York Times, 2015).

@@

The first well-known attempt to define accessibility quantitatively was by [@ingram1971concept], which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay. This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility]. The measurement of access to jobs by particular transport modes has received vast attention in the literature [@geurs2004], while studies on access to education has largely focused on the consequences of mode choice on socio-economic indicators and school outcomes (see for instance [@asahi2014], , [@falch2013] and [@dickerson2012] on the impact of accessibility on school outcomes, and [@andersson2012] and [@easton2015] on the effect of changes in travel-to-school patterns in developed economies).

The new index contributes to the measurement of school accessibility. It is inspired by the index developed by [@shen1998location] in the context of job accessibility. The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area. Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available. As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas. The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas. The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space. The inclusion of this competitive element to school accessibility, akin to that in job access, highlights the fact that under certain educational systems, the access to opportunities is mediated by competition.

Our work contributes to the literature on transport-related social exclusion in emerging economies [@jones2012] by providing a quantitative way to assess multiple dimensions into a single measure. It provides a benchmark to compare access inequalities in countries with unequal provision of public services [@gomide2006], [@pacione1989access]. In the context of contested mobilities, by capturing both the unequal distribution of schools and transport services across space, the index allows embedded inequalities to be better understood and, with political will, contested.

@@

```{r, echo=FALSE, fig.cap="The home location of students attending public secondary schools students in the SPMR.", out.width="70%"}
knitr::include_graphics(c("med_pu.png"))
```

```{r, echo=FALSE, fig.cap="The home location of students attending private secondary schools students in the SPMR.", out.width="70%"}
knitr::include_graphics(c("med_pr.png"))
```

@@

```{r, echo=FALSE, fig.cap="Enrollment in public secondary schools in the SPMR.", out.width="70%"}
knitr::include_graphics(c("enrol.png"))
```

@@

```{r table-placeholder, echo=FALSE, warning=FALSE}
tbl = read.csv("res-table.csv")
names(tbl) = c("Quantile", "Pre-reform", "Post-reform", "% change")
knitr::kable(tbl, row.names = F, caption = "Mean accessibility before and after redistribution")
```

@@

@@

 We conclude that extending the concept of local
accessibility indicators to education can help to both contest and
constructively tackle embedded social inequalities.
